---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Introduction

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment is available [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).

The variables included in this dataset are:

+ **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)
+ **date**: The date on which the measurement was taken in YYYY-MM-DD format
+ **interval**: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Loading and preprocessing the data

First load the data assuming the source file is in the current directory:
```{r}
activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
```
Next load the packages we will be using. We suppress the messages printed when lodaing these packages.
```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
```
Finally we preprocess the data.

First transform the date field into an R date format using the current time zone. Note that the `as.POSIXct` function knows how to handle dates formatted as in the spreadsheet, namely "2012-10-01".
```{r}
activity$date <- as.POSIXct(activity$date)
```
We look at the first few observations:
```{r}
head(activity)
```
The `interval` variable values appear to the number of minutes elapsed since midninght of the corresponding day in increment of $5$ minutes. However, the value following $55$ is $100$ not $60$. The format is a kind of numerical version of the `hours:minutes` format without the semicolon (and with all leading zeroes removed.) This is not very useful to plot our time series (see below) so we build a new variable that truly counts the number of minutes elapsed since midnight in $5$ minute increments:
```{r}
activity$minutesSinceMidnight <-
  floor(activity$interval/100) * 60 + activity$interval %% 100
```
For each day, this variable ranges from $0$ to $1435$ since there $1440$ minutes in a day.
```{r}
summary(activity$minutesSinceMidnight)
```

## What is mean total number of steps taken per day?

Note: we ignore the missing values in this part of the assignment

1. Compute the number of steps per day ignoring missing values:
```{r}
tot_steps_by_day <- activity %>%
  group_by(date) %>%
  summarise(totalSteps = sum(steps, na.rm = TRUE))
```

2. Make a histogram of the number of steps taken per day:
```{r, message=FALSE}
g <-ggplot(tot_steps_by_day, aes(x = totalSteps)) +
  geom_histogram(fill = "salmon", color = "black") +
  xlab("Total number of steps per day") + 
  ggtitle("Histogram of the total number of steps per day (missing data removed)")
g
```

3. Compute the mean and median values of the total number of steps taken per day:
```{r}
mean(tot_steps_by_day$totalSteps)
median(tot_steps_by_day$totalSteps)
```

## What is the average daily activity pattern?

1. Make a time series plot of the $5$-minute interval and the average number of steps taken within the interval, averaged across all days. First compute the averages:
```{r}
avge_by_interval <- activity %>%
  group_by(minutesSinceMidnight) %>%
  summarize(avgeSteps = mean(steps, na.rm = TRUE))
```
This dataset has $288$ rows, namely the number of $5$ minute intervals in a day.
```{r}
g = ggplot(avge_by_interval, aes(x=minutesSinceMidnight, y=avgeSteps)) +
  geom_line(color="steelblue") +
  xlab("Number of minutes elapsed since midnight") +
  ylab("Number of steps per 5 minute interval") +
  ggtitle("Average daily activity pattern for October and November 2012")
g
```

## Imputing missing values
1. We compute the total number of missing `steps` values:
```{r}
sum(is.na(activity$steps))
```
These missing values come from $8$ days that do not have any recorded `steps` values for a total of $8 * 288 = 2304$ values:
```{r}
dayNA = activity %>% 
  group_by(date) %>% 
  summarise(numNA = sum(is.na(steps))) %>% 
  filter(numNA > 0)
dayNA
```

2. We impute the missing values using the mean number of steps per $5$ minute interval across all days as computed in the previous section and rounded to the nearest integer.

3. We create a new dataset, `activityImputed`, that is equal to the original dataset but with the missing data filled in as described above. We then verify that there are no longer any missing `steps` values in the new dataset.
```{r}
activityImputed <- activity
for (dayIdx in dayNA$date)
  activityImputed[activityImputed$date == dayIdx,]$steps = round(avge_by_interval$avgeSteps,0)
sum(is.na(activityImputed$steps))
```

4. We compute the total number of steps taken each day from the imputed data:
```{r}
tot_steps_by_dayImputed <- activityImputed %>%
  group_by(date) %>%
  summarise(totalSteps = sum(steps, na.rm = TRUE))
```
And we make the corresponding histogram:
```{r, message=FALSE}
g <-ggplot(tot_steps_by_dayImputed, aes(x = totalSteps)) +
  geom_histogram(fill = "salmon", color = "black") +
  xlab("Total number of steps per day") + 
  ggtitle("Histogram of the total number of steps per day (using inputed data)")
g
```

We compute the new `mean` and `median` values of the total number of steps:
```{r}
mean(tot_steps_by_dayImputed$totalSteps)
median(tot_steps_by_dayImputed$totalSteps)
```
Observe that the new values for the `mean` and `median` are now essentially equal while the values obtained above when removing NA values were substantially different, with the value of the `mean` less than that of the `median`.

## Are there differences in activity patterns between weekdays and weekends?
1. We create a new factor in the dataset with two levels, `weekday` and `weekend` indicating whether a given date is a weekday or weekend day.
```{r}
activityImputed <- activityImputed %>%
  mutate(dayType = ifelse(weekdays(date) %in% c('Saturday','Sunday'), 'weekend', 'weekday'))
activityImputed$dayType <- factor(activityImputed$dayType)
```
2. We compute the average number of steps taken, averaged across all weekday days or weekend days.
```{r}
avge_by_intervalWeekday <- activityImputed %>%
  group_by(minutesSinceMidnight, dayType) %>%
  summarize(avgeSteps = mean(steps))
```
And we plot the corresponding time series for week days and for weekend says:
```{r}
g = ggplot(avge_by_intervalWeekday, aes(x=minutesSinceMidnight, y=avgeSteps)) +
  geom_line(color="steelblue") +
  facet_wrap(~ dayType, ncol = 1) +
  xlab("Number of minutes elapsed since midnight") +
  ylab("Number of steps per 5 minute interval") +
  ggtitle("Average daily activity pattern")
g
```

We observe definite differences in activity patterns between week days and weekend days. Activity starts earlier on week days (around 6:30 a.m.) and peaks higher around 9:00 a.m. However, we observe more activity on weekend days between mid-morning and about 8:30 p.m. than during week days, although the activity variation is higher on weekends than on week days.


